Simulating High-Dimensional Multivariate Data

### 2020-12-28 030-bigsimr-package.Rmd
4th draft. working on Ed comments

# The `bigsimr` R package

@@

Next, we specify the marginal distributions and correlation coefficient (both type and magnitude).
Here the analyst is free to be creative.
For this example, we do not take up goodness-of-fit considerations to determine the marginal distributions.
But it seems sensible without domain knowledge to estimate these quantities from the data and `bigsimr` contains fast functions designed for this task.

@@

To avoid manual entry, `mlist` presents an alternative to `alist` with on-the-fly evaluation.

@@

*Specifying correlation*.
As mentioned, the user must decide how to describe correlation, based on the particulars of the problem.
For non-normal data and for improved simulation accuracy in our scheme, we advocate the use of rank-based correlations Spearman's $\rho_S$ and Kendall's $\tau$.
But we also support approximate Pearson correlation coefficient matching, while cautioning the user to check the performance for their parametric multivariate model (see [Monte Carlo evaluations](#simulations) below for evaluation strategies and guidance).
To aid in correlation specification, and estimation in general, we provide a high-performance function `bigsimr::cor_fast` which estimates Pearson, Spearman, or Kendall correlation using the fastest methods available.
(Anyone who has tried estimating Kendall's $\tau$ using `stats::cor` can attest that the routine does not scale to even moderate dimensions).
These estimation methods are classical approaches, not designed for high-dimensional correlation estimation (see the [Conclusion and Discussion]({#discussion) sections for more on this).

```{r ch030-aq-cor, echo=TRUE}
type <- 'spearman'
(rho <- cor_fast(df, method = "spearman"))
```

*Checking the theoretical correlation bounds*.
As discussed in Section \@ref(background), given a pair of marginal distributions the possible correlations are not free to vary between $[-1, 1]$. 
`bigsimr::cor_bounds` estimates the theoretic lower and upper correlation bounds, using the Generate, Sort, and Correlate algorithm of @DH2011.

@@

Since our estimated Spearman correlation $\hat{ \rho}_S$ is within the theoretical bounds, the correlation is valid as input to `bigsimr:rvec`.
On the other hand, the bounds on the Pearson correlation coefficient $\rho_P$ between these margins is

```{r ch030-cor-bounds-pearson, echo=TRUE}
cor_bounds(margins = margins, type = 'pearson', reps = 1e6)
```

@@

```{r ch030-sim-margins, echo=TRUE}
x <- rvec(10000, rho, margins, type)
df_sim <- as.data.frame(x)
colnames(df_sim) <- colnames(df)
```

>>>>

# The `bigsimr` R package {#package}

@@

Next, we specify the marginal distributions and correlation coefficient (both type and magnitude).
Here the analyst is free to be creative.
For this example, we avoid goodness-of-fit considerations to determine the marginal distributions.
But it seems sensible without domain knowledge to estimate these quantities from the data and `bigsimr` contains fast functions designed for this task.

@@

To circumvent manual entry, `mlist` presents an alternative to `alist` with on-the-fly evaluation.

@@

*Specifying correlation*.
As mentioned, the user must decide how to describe correlation, based on the particulars of the problem.
For non-normal data and for improved simulation accuracy/scalability in our scheme, we advocate the use of Spearman correlation matrix $R_S$ and Kendall $\tau$ matrix $R_K$.
We also support approximate Pearson correlation coefficient matching, while cautioning the user to check the performance for their parametric multivariate model (see [Monte Carlo evaluations](#simulations) below for evaluation strategies and guidance).
To aid in correlation specification, and estimation in general, we provide a high-performance function `bigsimr::cor_fast` which estimates Spearman, Kendall, or Pearson correlation using the fastest methods available.
(Anyone who has tried estimating Kendall's $\tau$ using `stats::cor` can attest that the routine does not scale to even moderate dimensions).
These estimation methods are classical approaches, not designed for high-dimensional correlation estimation (see the [Conclusion and Discussion]({#discussion) sections for more on this).

```{r ch030-aq-cor, echo=TRUE}
type <- 'spearman'
(R_S <- cor_fast(df, method = "spearman"))
```

*Checking target correlation matrix admissibility*. First we use `bigsimr::cor_bounds` to ensure that the pairwise target correlation values are valid prior to mapping step which constructs $R_X$ for the MVN input into NORTA (see Section \@ref(algorithms)).
`cor_bounds` estimates the pairwise lower and upper correlation bounds using the Generate, Sort, and Correlate algorithm of @DH2011.

@@

Since our our single estimated Spearman correlation is within the theoretical bounds, the correlation is valid as input to `bigsimr:rvec`.
But even if the 2-dimensional bounds are satified for each pair of margins, this does not guarantee the feasibility of a $d-$variate distribution [@BF17].

To provide higher dimensional feasibility/admissibility checking, we begin by mapping using `bigsimr::cor_convert` and check admissibility.

```{r ch030-cor-convert, echo=TRUE}
## Step 1. Mapping
(R_X <- cor_convert(rho = R_S, from = type, to = 'pearson'))
## Step 2. Check admissibily
is_valid_correlation <- function(rho) {
  all(diag(rho) == 1) && matrixcalc::is.positive.semi.definite(rho)
}
is_valid_correlation( R_X )
matrixcalc::is.symmetric.matrix( R_X )
all(diag(R_X) == 1)
cor_nearPSD( G = R_X )
```

We note that the bounds on the Pearson correlation coefficient $R_P$ between these margins is restricted as seen below:

```{r ch030-cor-bounds-pearson, echo=TRUE}
cor_bounds(margins = margins, type = 'pearson', reps = 1e5)
```

@@

```{r ch030-sim-margins, echo=TRUE}
x <- rvec(10000, R_S, margins, type)
df_sim <- as.data.frame(x)
colnames(df_sim) <- colnames(df)
```

### 2020-12-28 050-examples.Rmd
4th draft. working on Ed comments

```{r ch050-estRhoBRCA, echo=TRUE, eval=TRUE, cache=F}
## Estimate Spearman's correlation on the count data
corType <- 'spearman'
nb_Rho <- bigsimr::cor_fast(brca, method = corType)
```

@@

>>>>

```{r ch050-estRhoBRCA, echo=TRUE, eval=TRUE, cache=F}
## Estimate Spearman's correlation on the count data
type <- 'spearman'
R_S <- bigsimr::cor_fast(brca, method = type)
```

@@

```{r ch050-cor-check, echo=TRUE, eval=FALSE, cache = FALSE}
## 1. Check 2D bounds
## OUR CODE HERE

## 2. Check admissibility
R_X <- cor_convert(rho = R_S, from = type, to = 'pearson')
## diag(R_X) <- 1
is_valid_correlation <- function(rho) {
  all(diag(rho) == 1) && matrixcalc::is.positive.semi.definite(rho)
}
is_valid_correlation( R_X )

system.time( R_Xtilde <- cor_nearPSD( G = R_X ) )
## diag(R_X) <- 1
is_valid_correlation( R_X )

norm( x = R_X, type = "F" )
norm( x = R_Xtilde, type = "F" )
norm( x = R_X - R_Xtilde, type = "F" )

```
